---
title: "Corona Assignment 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
```

### Data

- Download recent versions of the COVID-19 confirmed cases
    - from Johns Hopkins for the world
    - from *The New York Times* for the US
- Put the world data into long format
- Join longitude/latitude data to the US data set

```{r read-transform-data}

```


### Problems

1. Explain why some variable names have single back ticks around them, such as `3/26/20`.

2. How does `inner_join()` differ from `left_join()`, `right_join()`, and `full_join()`. Would it have made a difference in this situation which function to use?

3. Tabulate and graph the total number of confirmed cases in the world by date and graph this data.
Describe the trend.

```{r}

```

4. Make an animation of the US Map for total cases, similar to the world map animation.

```{r}

```

5. Add a new cases variable to the world data set (`group_by()` country, then `mutate` and use `diff()`) and make an animation of the world data showing new cases. This should show how hotspots are changing. Describe qualitatively what this animation shows about the COVID-19 global infection pattern over the past two months.

```{r}

```

