---
title: "Corona Assignment 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(lubridate)
install.packages("maps")
library(maps)
install.packages("mapproj")
library(mapproj)
install.packages("gganimate")
library(gganimate)
install.packages("gifski")
library(gifski)
```

### Data

- Download recent versions of the COVID-19 confirmed cases
    - from Johns Hopkins for the world
    - from *The New York Times* for the US
- Put the world data into long format
- Join longitude/latitude data to the US data set

```{r read-transform-data}
global = time_series_covid19_confirmed_global%>%
  pivot_longer(cols = 5:78, names_to = "Date", values_to = "Cases")

covid19 = time_series_covid19_confirmed_global %>% 
  mutate(`Country/Region` = case_when(
    `Province/State` == "Grand Princess" ~ "Grand Princess",
    `Province/State` == "Diamond Princess" ~ "Diamond Princess",
    TRUE ~ `Country/Region`
  )) %>% 
  filter(`Province/State` != "Recovered" | is.na(`Province/State`))
  

temp1 = covid19 %>% 
  select(`Country/Region`,starts_with("L"),`3/26/20`) %>% 
  group_by(`Country/Region`) %>% 
  summarize(Lat = sum(Lat*`3/26/20`) / sum(`3/26/20`),
            Long = sum(Long*`3/26/20`) / sum(`3/26/20`))

temp2 = covid19 %>% 
  select(-`Province/State`,-starts_with("L")) %>% 
  group_by(`Country/Region`) %>% 
  summarize_all(sum)


covid19_wide = inner_join(temp1,temp2) %>% 
  ungroup()


covid19_long = covid19_wide %>% 
  pivot_longer(cols = contains("/20"), names_to = "Date", values_to = "Total") %>% 
  mutate(Date = mdy(Date))
rm(temp1,temp2)


us_url = "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
us_covid19 = read_csv(us_url)

data(state)

state_centers = tibble(
  state = state.name,
  long = state.center$x,
  lat = state.center$y) %>% 
  bind_rows( tibble(
    state = "District of Columbia",
    long = -77.0369,
    lat = 38.9072))

us_covid19 = us_covid19 %>% 
  left_join(state_centers)
```


### Problems

1. Explain why some variable names have single back ticks around them, such as `3/26/20`.

1. Variable names that contain characters that are invalid for variable names. A foreword slash is an invalid character for a variable name, and a variable name cannot start with a number. Therefore, they must be contained within back ticks in order to pass as valid variable names. 

2. How does `inner_join()` differ from `left_join()`, `right_join()`, and `full_join()`. Would it have made a difference in this situation which function to use?

2. inner_join merges data frames by matching values, keeping all columns from both frames. left_join returns rows from one data frame, as well as columns from both. right_join does the same, in regards to the other data frame. full_join returns all rows and columns, with NA where matching data are missing. In regards to state longitudes and latitudes, if we had not used left_join, we would have gotten a lot of bulk in our frame, or we would have accidentally given state_centers the data from us_covid19. We use inner_join to rejoin the temporary data frames in order to avoid having multiple rows that contain data that could be stored in one, and also to have lat&long in the same row as cases by date.

3. Tabulate and graph the total number of confirmed cases in the world by date and graph this data.
Describe the trend.

3. The rate at which the diseased population grows seems to be exponential.
```{r}
p_3 = covid19_long%>%
  select(`Country/Region`, Date, Total)%>%
  group_by(Date)%>%
  mutate(world_total = sum(Total))%>%
  ungroup()%>%
  select(Date, world_total)

p_3

ggplot(data = p_3, mapping = aes(x = Date, y = world_total)) + geom_line()


```

4. Make an animation of the US Map for total cases, similar to the world map animation.

```{r}
us_map = map_data("state")

us_covid19 = us_covid19%>%
  filter(is.na(long)==FALSE, state!='Alaska', state!='Hawaii')

most_recent = us_covid19 %>% 
  pull(date) %>% 
  max()

## Filter us_covid19 to eliminate Alaska and Hawaii for the plots
## Use drop_na() to eliminate any data with no latitude or longitude


p = ggplot() +
  geom_polygon(data = us_map, aes(x=long,y=lat,group=group),
               fill="white",color="black") +
  geom_point(data = us_covid19,
             aes(x=long,y=lat,size=cases), alpha = 0.5, color="red") +
  scale_size_continuous(name="Cases", range=c(1,6)) +
  coord_map("bonne", lat0 = 50) +
  theme_void() + 
  guides( color = guide_legend() ) +
  labs( caption = "Data Repository provided by The New York Times.") +
  theme(
    legend.position = "bottom",
    text = element_text(color = "#22211d"),
    plot.background = element_rect(fill = "#ffffff", color = NA), 
    panel.background = element_rect(fill = "#ffffff", color = NA), 
    legend.background = element_rect(fill = "#ffffff", color = NA)
  )
p

p2 = p + transition_time(date) +
  labs(title = "Date: {frame_time}")

nframes = us_covid19%>%
  pull(date)%>%
  unique()%>%
  length()

gifski(animate(p2, nframes = nframes, fps = 5))

#I will submit the animation as well as this RMD
```

5. Add a new cases variable to the world data set (`group_by()` country, then `mutate` and use `diff()`) and make an animation of the world data showing new cases. This should show how hotspots are changing. Describe qualitatively what this animation shows about the COVID-19 global infection pattern over the past two months.

```{r}
p_5 = covid19_long%>%
  group_by(`Country/Region`)%>%
  mutate(new_cases = c(0,diff(Total)))


world = map_data("world")

mybreaks = c(1, 50, 100, 500, 1000, 5000, 10000, 50000, 100000)
mylabels = c("1-49", "50-99", "100-499", "500-999",
             "1,000-4,999", "5,000-9,999", "10,000-49,999",
             "50,000-99,999","100,000+")

g = ggplot() +
  geom_path(data = world, aes(x=long, y = lat, group = group), alpha=0.5) +  
  geom_point(data=filter(p_5, `Country/Region` != "Diamond Princess"),
             aes(x=Long, y=Lat, size=new_cases, color=new_cases),
             stroke=FALSE, alpha=0.25) +
  scale_size_continuous(name="New Cases", trans="log", range=c(1,6),
                        breaks=mybreaks, labels=mylabels) +
#  scale_color_viridis_c(option="inferno", name="Cases", trans="log",
#                        breaks=mybreaks, labels=mylabels) +
  scale_color_viridis_c(option="viridis", name="New Cases", trans="log",
                        breaks=mybreaks, labels=mylabels) +
  theme_void() + 
  guides( color = guide_legend() ) +
  labs( caption = "Data Repository provided by Johns Hopkins CSSE.") +
  theme(
    legend.position = "bottom",
    text = element_text(color = "#22211d"),
    plot.background = element_rect(fill = "#ffffff", color = NA), 
    panel.background = element_rect(fill = "#ffffff", color = NA), 
    legend.background = element_rect(fill = "#ffffff", color = NA)
  )

g

g2 = g + transition_time(Date) +
  labs(title = "Date: {frame_time}")

## count the number of dates to set nframes equal to that value
## x %>% pull(Date) is like x$Data
## unique() returns a vector of only the unique values
## length() is the length of the vector; here the number of unique dates
## we want one frame per date
nframes = covid19_long %>% 
  pull(Date) %>% 
  unique() %>% 
  length()

## Show to frames per second (fps = 2)
## This takes a couple minutes to render
gifski(animate(g2, nframes = nframes, fps = 10), gif_file = "animate2.gif")
```

